---
title: "A Step-by-Step Introduction to Structural Equation Modeling with lavaan in R"
subtitle: "Structural equation modelling in R with **lavaan**"
author: "George Michaelides"
institute: "School of Psychology, University of East Anglia"
date: "2026"
---

# Welcome {#welcome}

`lavaan-tutorial` is a tutorial site covering structural equation modelling (SEM) in R
using the **lavaan** package. It is written for MSc students on the **Statistical
Skills** module of the **MSc in Organisational Psychology** at the **University of
East Anglia**.

## Who this site is for

You are new to R, RStudio, and lavaan. You may have used SPSS before, but we will
not assume that — every concept is taught from the ground up in R/lavaan. By the
end of the 16 tutorials you will be able to specify, estimate, and report
confirmatory factor analyses, path models, structural equation models, and
mediation models in lavaan, and you will have a path into multilevel, growth, and
moderation analyses.

## Tutorials at a glance

```{r}
#| echo: false
#| fig-align: center
#| out-width: "90%"
DiagrammeR::grViz("
digraph {
  graph [rankdir = LR, fontsize = 10, fontname = Helvetica]
  node  [shape = box, style = filled, fillcolor = '#ecf0f1', fontname = Helvetica]

  F   [label = 'Foundations\\n1 - 4']
  B   [label = 'Bridging\\n5']
  C   [label = 'Core lavaan\\n6 - 12']
  Cap [label = 'Capstone\\n13']
  A   [label = 'Advanced\\n14 - 16']

  F -> B -> C -> Cap -> A
}
")
```

- **Foundations (1–4)** — R basics, RStudio, importing data, simple inference (t-test, ANOVA, lm).
- **Bridging (5)** — Introducing lavaan: declarative model specification.
- **Core lavaan (6–12)** — CFA, parcels and second-order factors, path analysis, SEM with latents, mediation.
- **Capstone (13)** — Reporting and exporting tables, figures, and APA write-ups.
- **Advanced (14–16)** — Multilevel models, growth curve models, moderation.

Start with **Tutorial 01** and work through in order; the later tutorials build on
the dataset and concepts introduced earlier. You can also jump to any tutorial
using the sidebar on the left, or use the search box at the top.

## Setup in five minutes

Install R from <https://cran.r-project.org/> and RStudio from
<https://posit.co/products/desktop/rstudio/>. Then, in the R console:

```{r}
#| eval: false
install.packages(c("haven", "lavaan", "ggplot2", "psych",
                   "tidySEM", "DiagrammeR", "readxl", "texreg"))
```

## How to use this site

Each tutorial follows the same structure:

1. **Learning objectives** — what you will be able to do by the end.
2. **Setup** — the package-loading code you need first; in later tutorials this appears in a collapsed `<details>` block.
3. **Concept** — narrative explanation, diagrams, and equations.
4. **Worked example** — full code, output, and interpretation paragraph.
5. **Common errors** — `::: {.callout-warning}` callouts flagging what tends to go wrong.
6. **Exercises** — inline `::: {.callout-tip}` Try-its during the tutorial; end-of-tutorial exercises with collapsed solutions in `::: {.callout-caution collapse="true"}`.
7. **Summary & next tutorial** — a one-paragraph bridge to the next page.
8. **Further reading** — academic papers, cited APA-style.

Six callout colours encode meaning throughout the site:

| Colour | Type | Meaning |
|--------|------|---------|
| Blue   | `note`      | Concepts and definitions |
| Green  | `tip`       | Inline Try-it exercises |
| Orange | `warning`   | Common errors to avoid |
| Red    | `important`  | Critical watch-outs |
| Yellow | `caution`   | Collapsed exercise solutions |

## Attribution

::: {.callout-note}
This lavaan tutorial series is a port of teaching materials originally developed
by **Kevin Daniels** (University of East Anglia) for his MSc Organisational
Psychology Statistical Skills module, which used Mplus. The R/lavaan adaptation
is by **George Michaelides** (University of East Anglia). The original Mplus
materials remain with their author.
:::

## Citation

If you use these tutorials in your teaching or research, please cite:

> Michaelides, G. (2026). *A step-by-step introduction to structural equation modeling with lavaan in R*. University of East Anglia. https://michaelides.github.io/lavaan-tutorial/

## Licence

Tutorial content is licensed under [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/).
Embedded R/lavaan code snippets are released under the MIT licence.

---
